---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# flexTab1

<!-- badges: start -->
<!-- badges: end -->

The goal of flexTab1 is to create a highly flexible Table 1 for descriptive statistics, allowing users to customize the output format, choose appropriate summary measures, and optionally display missing values, p-values (testing for difference between two groups), and standardized mean differences (SMD). It is designed to adapt to various reporting needs and ensures comprehensive data summaries in a user-friendly format.

This package is still an experimental version. 

## Installation

You can install the development version of flexTab1 from [GitHub](https://github.com/KGutmair/flexTab1)

``` r
# install.packages("pak")
pak::pak("KGutmair/flexTab1")
```

## Example

This is a basic example which shows you how to solve a common problem:

```{r example}
library(flexTab1)
## basic example code
```

