---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# flexTab1

<!-- badges: start -->
<!-- badges: end -->

The goal of flexTab1 is to create a highly flexible Table 1 for descriptive statistics, allowing users to customize the output format, choose appropriate summary measures, and optionally display missing values, p-values (testing for difference between two groups), and standardized mean differences (SMD). It is designed to adapt to various reporting needs and ensures comprehensive data summaries in a user-friendly format.

This package is still an experimental version. 

## Installation

You can install the development version of flexTab1 from [GitHub](https://github.com/KGutmair/flexTab1)

``` r
# install.packages("devtools")
devtools::install_github("KGutmair/flexTab1")
```

## Example

Table 1 for one group/treatment arm

```{r, echo = TRUE, message = FALSE, warning = FALSE}
library(dplyr)
library(flexTab1)


# Load pbc data from the survival package
pbc <- survival::pbc

baseline_var <- c("age", "chol", "sex", "stage", "platelet")

pbc <- pbc %>%
     mutate_at(c("stage", "trt", "edema", "hepato"), function(x) as.factor(x)) %>%
     filter(!is.na(trt)) %>%
     mutate(trt = ifelse(trt == "1", "D-penicillmain", "Placebo"))

 Table1_flex(
     data = pbc,
     variables = baseline_var 
   )


```

