---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)

```

# flexTab1

<!-- badges: start -->
An R package for creating flexible, publication-ready summary tables (Table 1) for descriptive statistics. Customize both statistical measures and table layout to meet your exact reporting needs.

[![R-CMD-check](https://github.com/KGutmair/flexTab1/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/KGutmair/flexTab1/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/KGutmair/flexTab1/graph/badge.svg)](https://app.codecov.io/gh/KGutmair/flexTab1)
<!-- badges: end -->


## Overview
flexTab1 creates highly flexible Table 1 summary tables for descriptive statistics. It allows you to customize both the statistical measures and the output layout. It supports:

-   **Statistical Flexibility**: Various summary measures can be selected for categorical (*absolute*, *relative*) and numerical variables (*mean*, *median*, *sd*, *min*, *max*, *quartile1*, *quartile3*), arranged in any combination. *Missing values* frequencies for each variable can be optionally displayed. For group comparisons, options include displaying *p-values* and *standardized mean differences* (SMD).

-   **Custom Layout**: The following layout options are available: *row sorting* and *column sorting*, the option to display *summary measure units* (e.g., n (%), median (SD)), and an empty line after each variable for improved readability. The output can be generated as either a *publication-ready* output table (created with the `flextable` package) or as a `data.frame`, allowing for further modifications or integration with other table formatting packages.

-   **Group Analysis**: Supports summary measures for a  *single group* and stratified by *two or more groups*. Additionally, `flexTab1` supports a nested group structure, allowing you to compare subgroups within main groups.

## Installation

You can install the development version of flexTab1 from [GitHub](https://github.com/KGutmair/flexTab1)

``` r
# install.packages("devtools")
devtools::install_github("KGutmair/flexTab1", build_vignettes = TRUE)
# Accessing the vignette
# vignette("Get_started", package = "flexTab1")
```

## Usage

### Loading data + preparation
```{r, echo = TRUE, message = FALSE, warning = FALSE, out.width='47%'}
library(dplyr)
library(flexTab1)
library(flextable)



# Load pbc data from the survival package
 pbc <- survival::pbc

   baseline_var <- c("age", "chol", "sex", "stage", "platelet")

   pbc[c("stage", "trt", "edema", "hepato")] <-
     lapply(pbc[c("stage", "trt", "edema", "hepato")], as.factor)

   pbc <- pbc[!is.na(pbc$trt), ]
 pbc$trt <- ifelse(pbc$trt == "1",
 "D-penicillamine",
 "Placebo")

```

### Table 1 for one group/treatment arm
```{r, out.width='40%'}

tab1_example <- Table1_flex(
     data = pbc,
     variables = baseline_var,
     measures_num = c("mean", "sd"),
     measures_cat = c("absolute", "relative")
   )

tab1_example %>%
   line_spacing(space = 1.7, part = "all") %>%
   autofit() 

```

### Table 1 for two treatment arms

```{r, echo = TRUE, message = FALSE, warning = FALSE, out.width='62%'}
tab1_example <- Table1_flex(
  data = pbc,
  variables = baseline_var,
  group_var = "trt")


tab1_example %>%
   line_spacing(space = 1.7, part = "all") %>%
   autofit() 
```



```{r, echo = TRUE, message = FALSE, warning = FALSE, out.width='62%'}
tab1_example <- Table1_flex(
  data = pbc,
  variables = baseline_var,
  group_var = "trt",
  add_measure_ident = FALSE,
  display_pvalue = TRUE, 
  display_smd = TRUE,
  sort_rows = c("age", "sex", "stage", "chol", "platelet"),
  measures_cat = c("absolute", "relative"),
  measures_num = c("median", "min", "max"),
  group_order = c("Placebo", "D-penicillamine")
)

tab1_example %>%
   line_spacing(space = 1.7, part = "all") %>%
   autofit() 
```





### Table 1 for nested groups

`flexTab1` supports a nested group structure, allowing you to compare subgroups within main groups.

```{r, echo = TRUE, message = FALSE, warning = FALSE, out.width='70%'}

baseline_var <- c("age", "chol", "platelet", "stage")
tab1_example <- Table1_flex(
  data = pbc,
  variables = baseline_var,
  group_var = "sex",
  treatment_arm = "trt",
  add_measure_ident = TRUE,
  display_pvalue = TRUE,
  measures_cat = c("relative"),
  measures_num = c("mean","sd"),
  flextable_output = TRUE
)

tab1_example %>%
   line_spacing(space = 1.7, part = "all") %>%
   autofit() 


```


