---
title: "Introduction"
output: rmarkdown::html_vignette
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{Introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(flexTab1)
```

# Introduction

This function creates a highly flexible summary measure table for descriptive statistics ("Table 1") with options to customize both the statistical options and the output layout. It supports following options:

-   **Group stratification:** Supports summary measured for a  *single group* (`group_var` = FALSE) and stratified by *two or more groups*. Additionally, a nested group structure is supported, allowing for the comparison of  subgroups (`group_var`) within main groups `treatment_arm`(e.g. compare the subgroups x and y within the main groups A and B)

-   **Statistical specifications:**

    -   `measure_cat`, `measure_num`: specifies, which summary measures should be displayed. For categorial variables, one can choose between the combination of "absolute" and "relative" frequency. For numerical variables, one can choose between any combination of "mean", "median", standard deviation "sd", minimum "min", maximum "max", first quartile "quartile1", third quartile "quartile3".

    -   `display_missings`: specifies, whether the absolute/relative frequencies of the missing values for each variable should be displayed.

    -   `display_pvalues`: specifies, whether the distribution of baseline variables between groups should be tested for differences. Its p-values are displayed in the output table. Until now, only a comparison between two groups is supported (for categorial variables: Chi-square test and Fisher\`s exact test, when expected cell count is freater than 5, for numerical variables: Wilcoxon test).

    -   `display_smd`: specifies, whether the standardized mean differences should be displayed. Currently, this supports only a two-group comparison.

-   **Layout specifications:**

    -   `new_line`: specifies, whether there should be an empty line after each variable

    -   `sort_rows`: specifies the row order, in which the variables should be displayed in the Table1.
        -   `group_order` and `treatment_order`: specifies the column order, in which the group categories should be displayed in the Table1.

    -   `add_measure_ident`: specifies, whether the summary measures identifiers (e.g. n (%), mean (SD) should be added to each variable.

    -   `flextable_output`: specifies, whether the function should return the Table 1 as `flextable` object, which can be instantly added to a htlm, word, latex or pdf-document. Another option included to return the Table 1 as data.frame allowing for further modifications or the applications of other tablular packages.
    

# Table 1 for different settings

Here, the Mayo Clinic Primary Biliary Cholangitis Data from the survival package will be used for creating a Table 1. 

We start with loading the required packages and the dataset in our repository. Furthermore, we get familiar with the variables in the data and doing some minor data cleaning necessary for the Table 1.
```{r, message=FALSE}
library("survival")
library(dplyr)

# Load pbc data from the survival package
pbc <- survival::pbc

baseline_var <- c("age", "chol", "stage", "platelet")

pbc[c("stage", "trt", "edema", "hepato")] <-
  lapply(pbc[c("stage", "trt", "edema", "hepato")], as.factor)

pbc <- pbc[!is.na(pbc$trt), ]
pbc$trt <- ifelse(pbc$trt == "1",
  "D-penicillamine",
  "Placebo"
)
```

The baseline_var vector contains all baseline variables for which we want to calculate descriptive summary measures. 

## Group stratifications
### One group
```{r, message=FALSE}
tab1_ex <- Table1_flex(
  data = pbc,
  variables = baseline_var
)
flextable::autofit(tab1_ex)
```

### Two and more groups
Usually, in a clinical or epidemiological trial, we want to compare baseline characteristics of two or multiple treatment groups/groups under exposure for possible imbalances. 
Here, we compare the baseline characteristics of the treatment group receiving D-penicillamine vs. placebo.

```{r, message=FALSE}
tab1_ex2 <- Table1_flex(
  data = pbc,
  variables = baseline_var,
  group_var = "trt",
  display_pvalue = TRUE
)
flextable::autofit(tab1_ex2)
```

One can also use the `treatment_arm` parameter to define the group variable. 

```{r, message=FALSE}
tab1_ex3 <- Table1_flex(
  data = pbc,
  variables = baseline_var,
  treatment_arm = "trt",
  display_pvalue = TRUE
)
flextable::autofit(tab1_ex3)
```


This stratification is, of course, not only limited to two groups. Note: In case of the comparison of more than two groups, there is no function implemented for the calculation of p-values and standardized mean differences yet.
```{r, message=FALSE}
tab1_ex4 <- Table1_flex(
  data = pbc,
  variables = c("age", "chol", "sex", "stage", "platelet"),
  group_var = "edema"
)
flextable::autofit(tab1_ex4)
```


### Nested groups
Sometimes, one wants to compare subgroups within main groups. In this example, we want to compare the baseline characteristics between the male and female group within each treatment group. 
```{r, message=FALSE}
tab1_ex5 <- Table1_flex(
  data = pbc,
  variables = c("age", "chol", "platelet", "stage"),
  group_var = "sex",
  treatment_arm = "trt",
  flextable_output = TRUE
)
flextable::autofit(tab1_ex5)
```



## Statistical specifications
### Measure summaries
The default setting for categorical variables is absolute and relative frequencies and for numerical variables the median and the minimum and maximum. One can customize those prespecified options, for example only realtive freuqencies for categorial measures and the 25% and 75% quantile (first and thirt quartile) for numerical measures. 

```{r, message=FALSE}
tab1_ex6 <- Table1_flex(
  data = pbc,
  variables = baseline_var,
  group_var = "trt",
  measures_cat = c("relative"),
  measures_num = c("median", "quantile1", "quantile3")
)
flextable::autofit(tab1_ex6)
```

Another example is:
```{r, message=FALSE}
tab1_ex7 <- Table1_flex(
  data = pbc,
  variables = baseline_var,
  group_var = "trt",
  measures_cat = c("relative", "absolute"),
  measures_num = c("mean", "sd")
)
flextable::autofit(tab1_ex7)
```

Note: till now, for numerical variables, up to three summary measures are allowed. 


### Display missing values 
By default, the absolute and relative frequencies of missing values are shown; but one can also opt this out: `display_missings`= FALSE.  
```{r, message=FALSE}
tab1_ex8 <- Table1_flex(
  data = pbc,
  variables = baseline_var,
  group_var = "trt",
  display_missings = FALSE
)
flextable::autofit(tab1_ex8)
```

```{r, message=FALSE}
tab1_ex8 <- Table1_flex(
  data = pbc,
  variables = baseline_var,
  group_var = "trt",
  display_missings = TRUE
)
flextable::autofit(tab1_ex8)
```

### Display p-values
Default is `display_pvalue` = FALSE. For numerical variables, differences in the distribution between two groups is done with a Wilcoxon test with an alpha level of 5\%. Differences in categorial variables are compared with a $\chi^2$ test or with a fisher exact test, if the number of expected counts in one cell of the contingency table exceeds 5.

```{r, message=FALSE}
tab1_ex9 <- Table1_flex(
  data = pbc,
  variables = baseline_var,
  group_var = "trt",
  display_pvalue = TRUE
)
flextable::autofit(tab1_ex9)
```


If `display_pvalue` = TRUE with a nested group structure, then testing for differences is done between subgroups (`group_var`) within the main group (`treatment_arm`)
```{r, message=FALSE}
tab1_ex10 <- Table1_flex(
  data = pbc,
  variables = c("age", "chol", "platelet", "stage"),
  group_var = "sex",
  treatment_arm = "trt",
  display_pvalue = TRUE
)
flextable::autofit(tab1_ex10)
```


### Display standardized mean differences (SMD)
Default is `display_smd` = FALSE. This only works for pairwise group comparisons. Compared to the p-value, which depends on sample size, the standardized mean difference (SMD) does not rely on formal hypothesis testing. For numerical variables, it standardizes the difference in group means by the pooled standard deviation, whereas for categorical variables, it standardizes the difference in group proportions (or frequencies) by the pooled variance of the proportions [@Austin31052011]. Since the mean differences are standardized, one can now better compare differences across different scales.Generally, a SMD between -0.05 and 0.05 can be considered as well balanced. 

```{r, message=FALSE}
tab1_ex11 <- Table1_flex(
  data = pbc,
  variables = baseline_var,
  group_var = "trt",
  display_pvalue = TRUE,
  display_smd = TRUE
)
flextable::autofit(tab1_ex11)
```

In nested groups: 
```{r, message=FALSE}
tab1_ex12 <- Table1_flex(
  data = pbc,
  variables = baseline_var,
  group_var = "sex",
  treatment_arm = "trt",
  display_pvalue = TRUE,
  display_smd = TRUE
)
flextable::autofit(tab1_ex12)
```


## Layout specifications
### New line
Adding a new line before a new variable for better readability 
```{r}
tab1_ex13 <- Table1_flex(
  data = pbc,
  variables = baseline_var,
  group_var = "sex",
  treatment_arm = "trt",
  display_smd = TRUE,
  new_line = TRUE
)
flextable::autofit(tab1_ex13)
```


### Row ordering
In default, all variables are sorted numerically. With the `sort_rows` parameter, one can customize this order in which the variables are displayed. 

```{r, message=FALSE}
tab1_ex14 <- Table1_flex(
  data = pbc,
  variables = baseline_var,
  group_var = "trt",
  display_pvalue = TRUE,
  sort_rows = c("stage", "age", "platelet", "chol")
)
flextable::autofit(tab1_ex14)
```


### Column ordering
```{r, message=FALSE}
tab1_ex15 <- Table1_flex(
  data = pbc,
  variables = baseline_var,
  group_var = "trt",
  display_pvalue = TRUE,
  sort_rows = c("stage", "age", "platelet", "chol"),
  group_order = c("Placebo", "D-penicillamine")
)
flextable::autofit(tab1_ex15)
```


```{r, message=FALSE}
tab1_ex15 <- Table1_flex(
  data = pbc,
  variables = baseline_var,
  group_var = "sex",
  treatment_arm = "trt",
  display_pvalue = TRUE,
  sort_rows = c("stage", "age", "platelet", "chol"),
  group_order = c("m", "f"),
  treatment_order = c("Placebo", "D-penicillamine")
)
flextable::autofit(tab1_ex15)
```


### Measure summary labelling

Default setting is `add_measure_ident`= TRUE
```{r, message=FALSE}
tab1_ex15 <- Table1_flex(
  data = pbc,
  variables = baseline_var,
  group_var = "sex",
  treatment_arm = "trt",
  display_pvalue = TRUE,
  sort_rows = c("stage", "age", "platelet", "chol"),
  group_order = c("m", "f"),
  treatment_order = c("Placebo", "D-penicillamine"),
  add_measure_ident = FALSE
)
flextable::autofit(tab1_ex15)
```


### Output options
Default setting is the output as a `flextable()`output, which returns a publications-ready format. This can be exported as html, pdf, world or inserted into a latex file. Alternatively, one can opt for returning the output as data.frame for additional modifications or another package for displaying the content as output table.  

Output as `data.frame`
```{r, message=FALSE}
out1 <- Table1_flex(
  data = pbc,
  variables = baseline_var,
  group_var = "trt",
  display_pvalue = TRUE,
  sort_rows = c("stage", "age", "platelet", "chol"),
  group_order = c("Placebo", "D-penicillamine"),
  flextable_output = FALSE
)
print(out1)
```

Output as `flextable` object + further modifications
```{r, message=FALSE}
out2 <- Table1_flex(
  data = pbc,
  variables = baseline_var,
  group_var = "trt",
  display_pvalue = TRUE,
  sort_rows = c("stage", "age", "platelet", "chol"),
  group_order = c("Placebo", "D-penicillamine"),
  flextable_output = TRUE
)

# Doing further adaptions using the flextable package
library(officer)
library(flextable)
big_border <- fp_border(color = "black", width = 3, style = "solid")
tab1_ex2 <- out2 %>%
  border_remove() %>%
  hline_top(part = "header", border = big_border) %>%
  hline_top(part = "body", border = big_border) %>%
  hline_bottom(part = "body", border = big_border) %>%
  fontsize(size = 9, part = "all")

flextable::autofit(tab1_ex2)
```


# References
